---
title: "Case Study 2: Employee Data Analysis"
author: "Meredith Ludlow & Kristen Rollins"
date: "December 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

# Summary

TODO (Meredith)

# Introduction

TODO (Kristen)

# Analysis

### Exploratory Data Analysis

```{r}
# Read in training data
dfTrain <- read.csv("CaseStudy2-data.csv")

# TODO maybe make categories for numeric vars like age, income, etc
```

If a variable did not have a significant impact on turnover, we would expect that the attrition percentage within a group is the same as the attrition percentage in the entire dataset. As we see below, in the whole training set 83.9% of employees stayed while 16.1% left. So, as we view the relative percentages for turnover for each categorical variable, those groups with high attrition percentages appear to have a strong impact on turnover. 

```{r}
# Percentage of retained/lost employees
table(dfTrain$Attrition) / nrow(dfTrain)

# Define categorical variables
cat_vars <- c("BusinessTravel", "Department", "Education", 
              "EducationField", "EnvironmentSatisfaction",
              "Gender", "JobInvolvement", "JobLevel",
              "JobRole", "JobSatisfaction", "MaritalStatus",
              "Over18", "OverTime", "PerformanceRating", 
              "RelationshipSatisfaction", "StockOptionLevel", 
              "WorkLifeBalance")

# TODO make tables look nicer
# Make relative frequency tables for categorical variables
for (var in cat_vars) {
  print(var)
  freqtable <- table(dfTrain[[var]], dfTrain$Attrition)
  count <- plyr::count(dfTrain[[var]])
  print(freqtable / count$freq)
}
```

```{r}
# TODO look at other variables

```

### Employee Trends

```{r}

```

### KNN Classification

```{r}
# Read in validation data
dfVal <- read.csv("CaseStudy2Validation.csv")

# Generate attrition predictions based on training data
#dfPreds <- ?

# Get accuracy of predictions
#confusionMatrix(table(dfVal$Attrition, dfPreds$Attrition))

# Write predictions to csv file
#write.csv(dfPreds, "CaseStudy2Predictions_Ludlow_Rollins.csv")
```

# Conclusion

TODO (Kristen)

# Presentation

This write-up is supplemented by video presentations from both Meredith and Kristen. The links are provided below.

Meredith: 

Kristen: 
