---
title: "Case Study 2: Employee Data Analysis"
author: "Meredith Ludlow & Kristen Rollins"
date: "December 9, 2018"
output:
  html_document:
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(knitr)
library(dplyr)
library(caret)
library(kableExtra)
```

# Summary

TODO 

# Introduction

The purpose of this analysis is to explore what variables are most associated with attrition levels in Fortune 1000 companies. Exploratory analytics will be used to determine these variables as well as other general trends associated with specific jobs. Finally, we will create a model that will predict whether or not an employee will leave the company voluntarily.

# Analysis

### Exploratory Data Analysis

```{r}
# Read in training data
dfTrain <- read.csv("CaseStudy2-data.csv")
```

If a variable did not have a significant impact on turnover, we would expect that the attrition percentage within a group is the same as the attrition percentage in the entire dataset. As we see below, in the whole training set 83.9% of employees stayed while 16.1% left. So, as we view the relative percentages for turnover for each categorical variable, those groups with high attrition percentages appear to have a strong impact on turnover. 

TODO Explanation of our varaible selection and which ones we left out.

```{r}
# Percentage of retained/lost employees
kable(table(dfTrain$Attrition) / nrow(dfTrain), 
      col.names=c("Attrition", "Percent")) %>% 
      kable_styling(full_width=FALSE)

# Define variables for analysis
variables <- c("BusinessTravel", "Department", "Education", 
              "EducationField", "EnvironmentSatisfaction",
              "Gender", "JobInvolvement", "JobLevel",
              "JobRole", "JobSatisfaction", "MaritalStatus",
              "OverTime", "PerformanceRating", 
              "RelationshipSatisfaction", "StockOptionLevel", 
              "WorkLifeBalance", "Age", "DailyRate", "DistanceFromHome", 
              "HourlyRate", "MonthlyIncome", "MonthlyRate", 
              "NumCompaniesWorked", "PercentSalaryHike",
              "TotalWorkingYears", "TrainingTimesLastYear", 
              "YearsAtCompany", "YearsInCurrentRole", 
              "YearsSinceLastPromotion", "YearsWithCurrManager")

# Turn numerical values into categorical
dfTrain$Age <- cut(dfTrain$Age, breaks=c(-Inf,30,40,50,Inf),labels=c(1,2,3,4))
dfTrain$DailyRate <- cut(dfTrain$DailyRate, breaks=c(-Inf,500,1000,Inf),labels=c(1,2,3))
dfTrain$DistanceFromHome <- cut(dfTrain$DistanceFromHome, breaks=c(-Inf,10,Inf),labels=c(1,2))
dfTrain$HourlyRate <- cut(dfTrain$HourlyRate, breaks=c(-Inf,65,Inf),labels=c(1,2))
dfTrain$MonthlyIncome <- cut(dfTrain$MonthlyIncome, breaks=c(-Inf,5000,10000,15000,Inf),labels=c(1,2,3,4))
dfTrain$MonthlyRate <- cut(dfTrain$MonthlyRate, breaks=c(-Inf,5000,10000,15000,20000,Inf),labels=c(1,2,3,4,5))
dfTrain$NumCompaniesWorked <- cut(dfTrain$NumCompaniesWorked, breaks=c(-Inf,4,Inf),labels=c(1,2))
dfTrain$PercentSalaryHike <- cut(dfTrain$PercentSalaryHike, breaks=c(-Inf, 13, 17, Inf), labels=c(1, 2, 3))
dfTrain$TotalWorkingYears <- cut(dfTrain$TotalWorkingYears, breaks=c(-Inf, 10, 20, Inf), labels=c(1, 2, 3))
dfTrain$TrainingTimesLastYear <- cut(dfTrain$TrainingTimesLastYear, breaks=c(-Inf, 1, 4, Inf), labels=c(1, 2, 3))
dfTrain$YearsAtCompany <- cut(dfTrain$YearsAtCompany, breaks=c(-Inf, 5, 15, 25, Inf), labels=c(1, 2, 3, 4))
dfTrain$YearsInCurrentRole <- cut(dfTrain$YearsInCurrentRole, breaks=c(-Inf, 5, 10, Inf), labels=c(1, 2, 3))
dfTrain$YearsSinceLastPromotion <- cut(dfTrain$YearsSinceLastPromotion, breaks=c(-Inf, 2, 7, Inf), labels=c(1, 2, 3))
dfTrain$YearsWithCurrManager <- cut(dfTrain$YearsWithCurrManager, breaks=c(-Inf, 5, 10, Inf), labels=c(1, 2, 3))
```

TODO discuss metric

```{r results="asis"}
# Make relative frequency tables for categorical variables
AbsDiff <- data.frame(Variable=character(), AvgDistance=numeric())
for (var in variables) {
  print(var)
  freqtable <- table(dfTrain[[var]], dfTrain$Attrition)
  count <- plyr::count(dfTrain[[var]])
  RelFreq <- freqtable / count$freq
  print(kable(RelFreq,row.names=TRUE) %>% kable_styling(full_width=FALSE))
  Sum <- sum(abs(RelFreq[,2]-0.1606838))/nrow(RelFreq)
  AbsDiff <- rbind(AbsDiff, data.frame(Variable=var, AverageDistance=Sum))
}
# Get average distance metric for all variables
AbsDiff <- AbsDiff[order(-AbsDiff$AverageDistance),]
kable(AbsDiff,row.names=FALSE) %>% kable_styling(full_width=FALSE)
```

### KNN Classification

```{r}
# Read in validation data
dfVal <- read.csv("CaseStudy2Validation.csv")

# Identify variables used to make predictions, based on avg dist metric
pred_vars <- c("OverTime", "JobRole", "JobInvolvement", "JobLevel", "MaritalStatus","WorkLifeBalance")

# Convert wanted factors into integers
dfTrain$OverTime <- as.integer(dfTrain$OverTime)
dfTrain$JobRole <- as.integer(dfTrain$JobRole)
dfTrain$MaritalStatus <- as.integer(dfTrain$MaritalStatus)
dfVal$OverTime <- as.integer(dfVal$OverTime)
dfVal$JobRole <- as.integer(dfVal$JobRole)
dfVal$MaritalStatus <- as.integer(dfVal$MaritalStatus)

# Generate attrition predictions based on training data
dfVal$dfPreds3 <- class::knn(dfTrain[,pred_vars], dfVal[,pred_vars], 
                             dfTrain$Attrition, k=3)
dfVal$dfPreds5 <- class::knn(dfTrain[,pred_vars], dfVal[,pred_vars], 
                             dfTrain$Attrition, k=5)

# Get accuracy of predictions
confusionMatrix(table(dfVal$Attrition, dfVal$dfPreds3))
confusionMatrix(table(dfVal$Attrition, dfVal$dfPreds5))

dfPreds <- select(dfVal, ID, dfPreds3)

# Write predictions to csv file
write.csv(dfPreds, "CaseStudy2Predictions_Ludlow_Rollins.csv")
```

### Employee Trends

```{r jobrole_jobsat, fig.align='center'}
# Re-load original data
dfTrain <- read.csv("CaseStudy2-data.csv")
# Job satisfaction by group
Jobs <- group_by(dfTrain, JobRole) %>% summarise(Avg=mean(JobSatisfaction, na.rm=TRUE))
kable(Jobs) %>% kable_styling(full_width = FALSE)

ggplot(data=Jobs, aes(x=JobRole, y=Avg, fill=JobRole)) + 
  geom_bar(stat='identity', colour = 'black') + 
  coord_flip() + 
  labs(title="Mean Job Satisfaction by Role", x="Job Role", y="Mean Satisfaction") +
  theme(legend.position="none") + 
  theme(plot.title = element_text(hjust = 0.5))
```

The average job satisfaction for each position is fairly close together. The lowest is Human Resources at 2.57 and the highest is Research Scientist and Healthcare Representative at 2.80. 

```{r income_jobsat}
# Scatterplot of Monthly Income by Job Role
ggplot(dfTrain, aes(x=JobRole, y=MonthlyIncome, group=JobRole)) + 
  ggtitle("Income Distribution by Job Role") + 
  xlab("Job Role") + 
  ylab("Monthly Income") + 
  geom_boxplot() +
  stat_summary(fun.y=mean, geom="point", colour="blue") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))
```

```{r}
# Scatterplot of Age by Job Role
ggplot(dfTrain, aes(x=JobRole, y=Age, group=JobRole)) + 
  ggtitle("Age Distribution by Job Type") + 
  xlab("Job Role") + 
  ylab("Age") + 
  geom_boxplot() +
  stat_summary(fun.y=mean, geom="point", colour="red") +
  theme(axis.text.x = element_text(angle = 50, hjust = 1))
```

```{r}
# Attrition rate by Job Role
freqtable <- table(dfTrain$JobRole, dfTrain$Attrition)
count <- plyr::count(dfTrain$JobRole)
RelFreq <- freqtable / count$freq
dfRel <- data.frame(RelFreq)
dfRel2 <- dfRel[10:18,]

ggplot(data=dfRel2, aes(x=Var1, y=Freq, fill=Var1)) + 
  geom_bar(stat='identity', colour = 'black') + 
  coord_flip() + 
  ggtitle("Attrition Rate per Job Type") + 
  ylab("Attrition Rate") + 
  xlab("Job Type") + 
  theme(legend.position="none") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Attrition rate by job involvement
freqtable <- table(dfTrain$JobInvolvement, dfTrain$Attrition)
count <- plyr::count(dfTrain$JobInvolvement)
RelFreq <- freqtable / count$freq
dfRel <- data.frame(RelFreq)
dfRel2 <- dfRel[5:8,]

ggplot(data=dfRel2, aes(x=Var1, y=Freq, fill=Var1)) + 
  geom_bar(stat='identity', colour = 'black') + 
  ggtitle("Attrition Rate by Job Involvement") + 
  ylab("Attrition Rate") + 
  xlab("Job Involvement") + 
  theme(legend.position="none") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Attrition rate by job level
freqtable <- table(dfTrain$JobLevel, dfTrain$Attrition)
count <- plyr::count(dfTrain$JobLevel)
RelFreq <- freqtable / count$freq
dfRel <- data.frame(RelFreq)
dfRel2 <- dfRel[6:10,]

ggplot(data=dfRel2, aes(x=Var1, y=Freq, fill=Var1)) + 
  geom_bar(stat='identity', colour = 'black') + 
  ggtitle("Attrition Rate by Job Level") + 
  ylab("Attrition Rate") + 
  xlab("Job Level") + 
  theme(legend.position="none") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Attrition rate by overtime
freqtable <- table(dfTrain$OverTime, dfTrain$Attrition)
count <- plyr::count(dfTrain$OverTime)
RelFreq <- freqtable / count$freq
dfRel <- data.frame(RelFreq)
dfRel2 <- dfRel[3:4,]

ggplot(data=dfRel2, aes(x=Var1, y=Freq, fill=Var1)) + 
  geom_bar(stat='identity', colour = 'black') + 
  ggtitle("Attrition Rate by Overtime") + 
  ylab("Attrition Rate") + 
  xlab("Overtime") + 
  theme(legend.position="none") + 
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}
# Attrition rate by marital status
freqtable <- table(dfTrain$MaritalStatus, dfTrain$Attrition)
count <- plyr::count(dfTrain$MaritalStatus)
RelFreq <- freqtable / count$freq
dfRel <- data.frame(RelFreq)
dfRel2 <- dfRel[4:6,]

ggplot(data=dfRel2, aes(x=Var1, y=Freq, fill=Var1)) + 
  geom_bar(stat='identity', colour = 'black') + 
  ggtitle("Attrition Rate by Job Involvement") + 
  ylab("Attrition Rate") + 
  xlab("Job Involvement") + 
  theme(legend.position="none") + 
  theme(plot.title = element_text(hjust = 0.5))
```
# Conclusion

TODO

# Presentation

This write-up is supplemented by video presentations from both Meredith and Kristen. The links are provided below.

Meredith: 

Kristen: 
